Integrated nested Laplace approximations for extended latent Gaussian models with application to the Naomi HIV model

### 2023-02-10 src/docs_paper/paper.Rmd
Import KS plot and table into main text

@@

Consider a country in sub-Saharan Africa where a household survey with complex survey design has taken place during at time $T_1$.
Let $x$ index district, $a$ five-year age band, and $s$ sex.
For ease of notation, let $i$ index the finest district-age-sex division included in the model.
The data we observe may be aggregated over indices $i$, so we let $\mathcal{I}$ be a set of $i$ for which observations are reported.
Let $N_i \in \mathbb{N}$ be the population size, $\rho_i \in [0, 1]$ be HIV prevalence, $\alpha_i \in [0, 1]$ be antiretroviral therapy (ART) coverage, and $\lambda_i > 0$ be the annual HIV incidence rate.
We specify independent mixed effects models for HIV prevalence and ART coverage in the general population on the logit scale such that
\begin{align*}
\text{logit}(\rho_i) &= \eta^\rho_i, \\
\text{logit}(\alpha_i) &= \eta^\alpha_i,
\end{align*}
for certain choice of linear predictors $\eta^\rho_i$ and $\eta^\alpha_i$.
For the HIV incidence rate we use a mixed effects model on the log scale
$$
\log(\lambda_i) = \eta^\lambda_i(\{\rho_i, \alpha_i\}_{i \in \mathcal{I}}),
$$
where the linear predictor depends on $\{\rho_i, \alpha_i\}_{i \in \mathcal{I}}$ for some $\mathcal{I}$.
Let $\kappa_i$ be the proportion recently infected among HIV positive persons.
For each set of observed strata indices $\mathcal{I}$, we calculate the weighted observations $\hat \theta_\mathcal{I}$ for $\theta \in \{\rho, \alpha, \kappa\}$ with respective Kish effective sample sizes 
$$
M^{\hat \theta}_\mathcal{I} = \frac{\left(\sum_j w_j\right)^2}{\sum_j w_j^2},
$$
where $j$ indexes individuals across all strata $i \in \mathcal{I}$, with corresponding survey weights $w_j$.
The observed number of indicator cases is then
$$
Y^{\hat \theta}_{\mathcal{I}} = M^{\hat \theta}_{\mathcal{I}} \cdot \hat \theta_{\mathcal{I}}.
$$
For $\theta \in \{\rho, \alpha, \kappa\}$ we model these aggregate observations using a binomial working likelihood
$$
Y^{\hat \theta}_{\mathcal{I}} \sim \text{xBin}(M^{\hat \theta}_{\mathcal{I}}, \theta_{\mathcal{I}}),
$$
where $\theta_{\mathcal{I}}$ are the following appropriately weighted aggregates
\begin{align*}
\rho_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i}{\sum_{i \in \mathcal{I}} N_i}, \\
\alpha_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i \alpha_i}{\sum_{i \in \mathcal{I}} N_i \rho_i}, \\
\kappa_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i \kappa_i}{\sum_{i \in \mathcal{I}} N_i \rho_i}.
\end{align*}
To link the proportion recently infected among HIV positive persons $\kappa_i$ to HIV incidence $\lambda_i$ we use
$$
\kappa_i = 1- \exp \left( - \lambda \cdot \frac{1 - \rho_i}{\rho_i} \cdot (\Omega_T - \beta_T) - \beta_T \right),
$$
where the mean duration of recent infection $\Omega_T$ and false recent ratio $\beta_T$ are strongly informed by priors for the particular survey.

@@

Let $\dot A_\mathcal{I}$ be the number of people receiving ART.

@@

Using a common `TMB` template, we fit the simplified Naomi model to data from Malawi using four inferential approaches: (1) EB combined with a Gaussian approximation using `TMB`, (2) AGHQ combined with a Gaussian approximation using `aghq`, (3) AGHQ combined with a Laplace approximation by extending `aghq` and (4) the Hamiltonian Monte Carlo (HMC) algorithm No-U-Turn Sampling (NUTS) using `tmbstan`.
Details regarding particular settings used for each inferential method are provided in Table \ref{tab:inference-methods}.
The \textsc{R} [@r] code used to produce all results we describe is available at `github.com/athowes/elgm-inf`.
We used `orderly` [@orderly] for reproducible research, `ggplot2` for data visualisation [@wickham2016ggplot2] and `rticles` [@allaire2022rticles] for reporting via `rmarkdown` [@allaire2022rmarkdown].

\begin{table}[]
\small
\begin{tabularx}{\textwidth}{p{0.2\linewidth}p{0.75\linewidth}}
\toprule
Inferential method & Details \\
\midrule
1. EB, Gaussian & 0, $1000$ samples \\
2. AGHQ, Gaussian & $k = 1$, $1000$ samples \\
3. AGHQ, Laplace & $k = 1$, $l = 5$, $1000$ samples \\
4. NUTS & $4$ chains of $20000$ iterations with the first $10000$ iterations of each chain discarded as warmup, then thinned by a factor of $20$. HMC parameters set to default for \texttt{rstan}. \\
\bottomrule
\end{tabularx}
\caption{A summary of settings used for each inferential method.}
\label{tab:inference-methods}
\small
\end{table}

@@

>>>>

```{r echo = FALSE}
options(scipen = 100)
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  dpi = 320,
  cache = TRUE,
  out.width = "95%",
  fig.align = 'center'
)
```

@@

Consider a country in sub-Saharan Africa where a household survey with complex survey design has taken place at time $T_1$.
Let $x$ index district, $a$ five-year age band, and $s$ sex.
For ease of notation, let $i$ index the finest district-age-sex division included in the model.
The data we observe may be aggregated over indices $i$, so we let $\mathcal{I}$ be a set of $i$ for which observations are reported.
Let $N_i \in \mathbb{N}$ be the population size, $\rho_i \in [0, 1]$ be HIV prevalence, $\alpha_i \in [0, 1]$ be antiretroviral therapy (ART) coverage, and $\lambda_i > 0$ be the annual HIV incidence rate.
Independent mixed effects models for HIV prevalence and ART coverage in the general population are specified on the logit scale such that
\begin{align*}
\text{logit}(\rho_i) &= \eta^\rho_i, \\
\text{logit}(\alpha_i) &= \eta^\alpha_i,
\end{align*}
for certain choice of linear predictors $\eta^\rho_i$ and $\eta^\alpha_i$.
For the HIV incidence rate we use a mixed effects model on the log scale $\log(\lambda_i) = \eta^\lambda_i(\{\rho_i, \alpha_i\}_{i \in \mathcal{I}})$, where the linear predictor depends on $\{\rho_i, \alpha_i\}_{i \in \mathcal{I}}$ for some $\mathcal{I}$.
Let $\kappa_i$ be the proportion recently infected among HIV positive persons.
For each set of observed strata indices $\mathcal{I}$, we calculate the weighted observations $\hat \theta_\mathcal{I}$ for $\theta \in \{\rho, \alpha, \kappa\}$ with respective Kish effective sample sizes 
$$
M^{\hat \theta}_\mathcal{I} = \frac{\left(\sum_j w_j\right)^2}{\sum_j w_j^2},
$$
where $j$ indexes individuals across all strata $i \in \mathcal{I}$, with corresponding survey weights $w_j$.
The observed number of indicator cases is then $Y^{\hat \theta}_{\mathcal{I}} = M^{\hat \theta}_{\mathcal{I}} \cdot \hat \theta_{\mathcal{I}}$.
For $\theta \in \{\rho, \alpha, \kappa\}$ we model these aggregate observations using a binomial working likelihood $Y^{\hat \theta}_{\mathcal{I}} \sim \text{xBin}(M^{\hat \theta}_{\mathcal{I}}, \theta_{\mathcal{I}})$, where $\theta_{\mathcal{I}}$ are the following appropriately weighted aggregates
\begin{align*}
\rho_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i}{\sum_{i \in \mathcal{I}} N_i}, \\
\alpha_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i \alpha_i}{\sum_{i \in \mathcal{I}} N_i \rho_i}, \\
\kappa_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i \kappa_i}{\sum_{i \in \mathcal{I}} N_i \rho_i}.
\end{align*}
To link the proportion recently infected among HIV positive persons $\kappa_i$ to HIV incidence $\lambda_i$ we use
$$
\kappa_i = 1- \exp \left( - \lambda \cdot \frac{1 - \rho_i}{\rho_i} \cdot (\Omega_T - \beta_T) - \beta_T \right),
$$
where the mean duration of recent infection $\Omega_T$ and false recent ratio $\beta_T$ are strongly informed by priors for the particular survey.

@@

Let $\dot A_\mathcal{I} = \sum_{x \sim x', x = x'} \dot A_{x', x}$ be the number of people receiving ART.
Model $\dot A_\mathcal{I} \sim \mathcal{N}(\tilde A_\mathcal{I}, \sigma^{\tilde A_\mathcal{I}})$ using a normal distribution with mean $\sum_{x \sim x', x = x'} N_{x'} \pi_{x', x}$ and standard deviation $\sigma^{\tilde A_\mathcal{I}}) = \sqrt{\sum_{x \sim x', x = x'} N_{x'} \pi_{x', x} (1 - \pi_{x', x})}$.

@@

Using one `TMB` template (available in the appendix), we fit the simplified Naomi model to data from Malawi using four inferential approaches: (1) EB combined with a Gaussian approximation using `TMB`, (2) AGHQ combined with a Gaussian approximation using `aghq`, (3) AGHQ combined with a Laplace approximation by extending `aghq` and (4) the Hamiltonian Monte Carlo (HMC) algorithm No-U-Turn Sampling (NUTS) using `tmbstan`.
Details regarding particular settings used for each inferential method are provided in Table \ref{tab:inference-methods}.

The \textsc{R} [@r] code used to produce all results we describe below is available at `github.com/athowes/elgm-inf`.
We used `orderly` [@orderly] for reproducible research, `ggplot2` for data visualisation [@wickham2016ggplot2] and `rticles` [@allaire2022rticles] for reporting via `rmarkdown` [@allaire2022rmarkdown].

\begin{table}[]
\small
\begin{tabularx}{\textwidth}{p{0.2\linewidth}p{0.75\linewidth}}
\toprule
Inferential method & Details \\
\midrule
1. EB, Gaussian & $1000$ samples \\
2. AGHQ, Gaussian & $k = 1$, $1000$ samples \\
3. AGHQ, Laplace & $k = 1$, $l = 5$, $1000$ samples \\
4. NUTS & $4$ chains of $20000$ iterations with the first $10000$ iterations of each chain discarded as warmup, then thinned by a factor of $20$. HMC parameters set to default for \texttt{rstan}. \\
\bottomrule
\end{tabularx}
\caption{A summary of settings used for each inferential method.}
\label{tab:inference-methods}
\small
\end{table}

@@

```{r}
knitr::include_graphics("depends/ks_summary_tmb_adam.png")
```

```{r}
ks_summary <- readRDS("depends/ks_summary_out.rds")

gt::gt(ks_summary)
```
